Multi-level modeling with Openflexo/\FML

### 2022-03-09 2-abstract.tex
Updates from Overleaf

Model federation is a multi-model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links (which are not level bounded) serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{MULTI process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Flexo Modeling Language ({FML}) that serves to describe them as the main building block in order to solve the process modeling challenge whereas the federation aspect is used as a means to provide editing tooling for the resulting process language. Our solution is fully implemented with the Openflexo framework. It fulfills all the challenge requirements.

>>>>

Model federation is a multi-model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links (which are not level bounded) serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{MULTI process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Federation Modeling Language ({FML}) that serves to describe them as the main building block in order to solve the process modeling challenge whereas the federation feature {\color{red} Do you have a better word than aspect here?} is used as a means to provide editing tools {\color{red} I would rather use "tools" instead of "tooling".  Tooling is the process of adding/making tools and you are already using "provide" before the "editing tooling"} for the resulting process language. Our solution fulfills all the challenge requirements and is fully implemented with the Openflexo framework. {\color{red} For the list of keywords, nobody will search your article using them :), use generic search keywords that point to your article. }

### 2022-03-09 4-technology.tex
convert drawio pdf to 1.5
fix spaces
Updates from Overleaf

To meet the \mlpc, we have decided to use the language infrastructure of
model federation approach~\parencite{Golra2016-federation}. \emph{Model
  federation} is a way to assemble models using some kind of
low-coupling links. It has been studied to answer the \OMG' RFP
Semantic Modeling for Information Federation~\parencite{simf}. In
contrast to approaches that compose metamodels into a single large
metamodel grouping all needed entities, model federation build links
between models and metamodels (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/metamodel conformity -- that is presented
in~\parencite{models2016-freemodel}. Another notable feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.
So to provide an adaptable framework, we have developed a flexible modeling language to provide modeling features without strict modeling levels. In a federation, the targeted modeling approaches can be unknown at the beginning and also can integrate several modeling levels.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.0 \textwidth]{Figures/FMLMetaModel.pdf}
    \caption{Representation of the main concepts of {FML} language}
    \label{fig:mm}
\end{figure*}
%\noteAntoine{Dans la figure ne manque-t-il pas les containements ? (VM vers Concepts)}
%\noteSylvain{C'est le cas dans cette nouvelle figure}

This framework relies on the architecture of Fig.~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Federation Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

A \UML like representation of a simplified metamodel of \FML is provided in Fig.~\ref{fig:mm}. \FML is a language designed to define models as virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \textsf{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{Figures/BPMNSubsetExample.pdf}
    \caption{An example of \BPMN subset showing both the model and an instance of that model}
    \label{fig:BPMNSubsetExample}
\end{figure}

>>>>

To meet the \mlpc, we have decided to use the language infrastructure of
model federation approach~\parencite{Golra2016-federation}. \emph{Model
  federation} is a way to assemble models using some kind of
low-coupling links. It was initially  formulated to respond to the \OMG's RFP
Semantic Modeling for Information Federation~\parencite{simf}. In
contrast to approaches that compose metamodels into a single large
metamodel grouping all needed entities, model federation build links
between models and metamodels (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/metamodel conformity -- that is presented
in~\parencite{models2016-freemodel}. Another notable feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.
So to provide an adaptable framework, we have developed a flexible modeling language to provide modeling features without strict modeling levels. In a federation, the targeted modeling approaches can be unknown at the beginning and also can integrate several modeling levels.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.0 \textwidth]{Figures/FMLMetaModel-1.5.pdf}
    \caption{Representation of the main concepts of {FML} language}
    \label{fig:mm}
\end{figure*}
%\noteAntoine{Dans la figure ne manque-t-il pas les containements ? (VM vers Concepts)}
%\noteSylvain{C'est le cas dans cette nouvelle figure}

This framework relies on the architecture of Fig.~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Federation Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling {\color{red} tooling is used in place of tool throughout the article.  I suggest clarifying it and then editing it all along the article}. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML, its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

{\color{red} This paragraph needs a parallel in naming convention between Figure 2 and the text.  FlexoConcept, FlexoBehavior, etc.} A \UML like representation of a simplified metamodel of \FML is provided in Fig.~\ref{fig:mm}. \FML is a language designed to define virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \textsf{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{Figures/BPMNSubsetExample-1.5.pdf}
    \caption{An example of \BPMN subset showing both the model and an instance of that model}
    \label{fig:BPMNSubsetExample}
\end{figure}
